---
title: "Bias Mitigation Pipeline in AI-Driven Hiring Processes: A Three-Dataset R-Based DALC/MLOps Approach"
author: ""
output: rticles::ieee_article
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(knitr)
library(kableExtra)
library(dplyr)
library(readr)
```

# Abstract

Artificial intelligence systems deployed in hiring pipelines risk perpetuating historical biases related to gender, ethnicity, and socioeconomic background. This paper presents an end-to-end bias mitigation pipeline for AI-driven hiring, implemented in R following the Data Analytics Life Cycle (DALC) and MLOps methodology. The pipeline covers six phases: data preparation, exploratory data analysis with bias detection, modeling with reweighting-based mitigation, fairness evaluation, explainable AI (XAI) analysis, and deployment via both a Plumber REST API and a Shiny dashboard. We validate our approach across three datasets: a synthetic hiring dataset (2,000 applicants, 7 features), a Kaggle AI Hiring Bias Detection dataset (1,500 candidates, 11 features), and the FairJob dataset (up to 2,000 candidates, 9 features). Four classification models are trained per dataset: a baseline Random Forest and three reweighted variants (Random Forest, Naive Bayes, and K-Nearest Neighbors). Data is split 70/15/15 into training, validation, and test sets. Fairness is assessed using demographic parity, equalized odds, calibration, and disparate impact ratio, with gender as the protected attribute. Explainability is provided through DALEX-based feature importance, break-down attribution, and partial dependence profiles. On Dataset 3 (FairJob), the mitigated Naive Bayes achieves a 73.5% reduction in demographic parity difference. On Dataset 2, the mitigated Random Forest achieves 0.894 accuracy and 0.916 AUC while improving fairness. The three-dataset validation demonstrates that a single reweighting-based pipeline generalizes across datasets with different characteristics.

*Keywords --- bias mitigation, fairness, AI hiring, reweighting, explainable AI, DALEX, DALC, MLOps, R, calibration*

# I. Introduction

The integration of artificial intelligence into recruitment and hiring processes has become widespread, with organizations deploying machine learning models to screen resumes, rank candidates, and make hiring recommendations [1]. While these systems promise efficiency and consistency, they also risk encoding and amplifying historical biases present in training data [2]. The well-documented case of Amazon's hiring tool, which systematically penalized female candidates, illustrates the real-world consequences of deploying biased AI systems in high-stakes decision contexts [4].

Addressing algorithmic bias in hiring requires a systematic approach that encompasses bias detection, mitigation, evaluation, and explainability [5]. The fairness-aware machine learning literature has produced a rich toolkit of metrics and intervention strategies [6], yet practical implementations that demonstrate end-to-end pipelines across multiple datasets remain scarce.

This paper contributes a comprehensive, reproducible bias mitigation pipeline built in R, following the Data Analytics Life Cycle (DALC) and MLOps best practices. Our pipeline is validated across three distinct datasets to address three research questions:

- **RQ1**: Does the baseline model exhibit measurable gender bias in hiring predictions?
- **RQ2**: Can reweighting-based mitigation reduce demographic parity difference while maintaining acceptable predictive performance?
- **RQ3**: Does the pipeline generalize across datasets with different characteristics, structures, and bias profiles?

The remainder of this paper is organized as follows: Section II reviews related work. Section III details our methodology. Section IV describes the three datasets. Section V presents experimental results. Section VI provides XAI analysis. Section VII discusses implications. Section VIII addresses limitations, and Section IX concludes.

# II. Literature Review

## A. Algorithmic Fairness Foundations

Barocas and Hardt [5] provide a comprehensive treatment of fairness definitions and their mathematical relationships. Chouldechova [8] formally proved that calibration and error rate balance cannot simultaneously hold except in degenerate cases. Dwork et al. [11] introduced individual fairness, requiring similar individuals to receive similar outcomes. Corbett-Davies and Goel [9] survey fairness criteria and argue for cost-sensitive approaches.

## B. Bias Mitigation Strategies

Bias mitigation techniques are categorized into pre-processing, in-processing, and post-processing approaches [6]. Kamiran and Calders [10] pioneered reweighting as a pre-processing technique, assigning inverse-frequency weights based on protected group and outcome class membership. Bellamy et al. [7] implemented reweighting in the IBM AI Fairness 360 toolkit. Friedler et al. [12] benchmarked fairness interventions across multiple datasets, finding no single technique dominates across all settings.

## C. AI Hiring Bias

The 2018 Reuters report on Amazon's biased hiring tool [4] demonstrated how historical gender imbalances in training data led to systematic discrimination. Recent IEEE work [1] has addressed bias mitigation in hiring using explainable AI techniques. The 2025 IEEE study on ethical challenges in resume screening [2] highlights the importance of fairness-oriented approaches. The DLA Piper 2025 AI framework [3] establishes new standards for tackling algorithmic bias in organizational contexts.

## D. Explainable AI for Fairness

DALEX provides model-agnostic tools for feature importance, break-down attribution, and partial dependence profiles [7]. The fairmodels R package extends DALEX with fairness checking capabilities, enabling integrated fairness-explainability workflows.

# III. Methodology

Our pipeline follows the six-phase Data Analytics Life Cycle (DALC), adapted for MLOps deployment.

## A. Phase 1: Data Preparation

Data preparation involves loading, cleaning, and splitting each dataset. Categorical variables are encoded as factors, identifier columns are removed to prevent data leakage, and missing values are handled through complete-case filtering.

All three datasets undergo a **70/15/15 stratified train-validation-test split** on the outcome variable using a fixed random seed (1266). The validation set is used for model selection; the test set is reserved for final evaluation.

## B. Phase 2: EDA and Bias Detection

EDA encompasses three analyses per dataset: (1) hiring/selection rates by gender, (2) feature distribution comparisons across gender groups, and (3) correlation analysis. The disparate impact ratio is computed as $DI = \frac{\min(rate_{g_0}, rate_{g_1})}{\max(rate_{g_0}, rate_{g_1})}$.

## C. Phase 3: Modeling with Reweighting

We train four models per dataset: a **baseline Random Forest** (no mitigation) and three reweighted variants using **Random Forest**, **Naive Bayes**, and **K-Nearest Neighbors (KNN)**.

Reweighting follows the inverse-frequency approach of Kamiran and Calders [10]:

$$w_{g,y} = \frac{N}{|G| \times n_{g,y}}$$

For Random Forest, weights are applied as case weights via the tidymodels workflow framework. For Naive Bayes and KNN, which do not natively support case weights, weighted resampling is used: training instances are sampled with replacement proportional to their computed weights, producing a rebalanced training set.

All models use a standardized recipe with dummy encoding of nominal predictors, zero-variance feature removal, and numeric normalization.

## D. Phase 4: Fairness Evaluation

Four fairness metrics are computed per model:

- **Demographic Parity Difference (DPD)**: $|P(\hat{Y}=1|G=1) - P(\hat{Y}=1|G=0)|$.

- **Equalized Odds Difference (EOD)**: $\max(|TPR_1 - TPR_0|, |FPR_1 - FPR_0|)$.

- **Calibration Difference**: $|(\bar{p}_1 - \bar{y}_1) - (\bar{p}_0 - \bar{y}_0)|$, where $\bar{p}_g$ is the mean predicted probability and $\bar{y}_g$ is the actual positive rate for group $g$. This measures whether the model is equally well-calibrated across gender groups.

- **Disparate Impact Ratio (DIR)**: $\frac{\min(rate_0, rate_1)}{\max(rate_0, rate_1)}$.

Bias reduction percentage is calculated as $\frac{DPD_{baseline} - DPD_{best\_mitigated}}{DPD_{baseline}} \times 100$.

## E. Phase 5: Explainability (XAI)

We employ DALEX-based tools: permutation-based feature importance, break-down attribution for individual predictions, partial dependence profiles (PDP), and gender-stratified ROC curves.

## F. Phase 6: Deployment

The best model per dataset is deployed via a **Plumber REST API** with endpoints `/health`, `/predict`, and `/fairness_report`. A **Shiny dashboard** provides an interactive interface for HR teams, displaying model metrics, fairness evaluation, before/after comparison, and a prediction form for all three datasets.

# IV. Dataset Description

## A. Dataset 1: Synthetic Hiring Dataset

Dataset 1 (`final_bias_hiring_dataset.csv`) contains 2,000 applicant records. After dropping identifiers, 7 features are retained. The dataset is gender-balanced (991 female, 1,009 male) with a 57.8% selection rate and a disparate impact ratio of 0.980.

## B. Dataset 2: Kaggle AI Hiring Bias Detection

Dataset 2 (`data.csv`) contains 1,500 candidate records with 11 features. It contains 762 female and 738 male candidates. The hiring rate is 31.0% with a disparate impact ratio of 0.993, representing near-perfect gender parity.

## C. Dataset 3: FairJob Dataset

Dataset 3 (`job fair dataset.csv`) is sourced from the FairJob challenge and contains employment outcome data with 9 features after preprocessing. Records are filtered to include only binary gender (Man/Woman), and a random sample of up to 2,000 is taken. Features include Age (categorical), Education Level, Employment status, Mental Health, Years of Coding experience, Previous Salary, and Computer Skills. The target variable is `Employed` (0/1).

## D. Dataset Comparison

```{r table3, results='asis'}
comparison <- data.frame(
  Characteristic = c("Records", "Features (modeled)", "Positive outcome rate",
                      "Disparate impact ratio", "Source", "Target variable"),
  Dataset1 = c("2,000", "7", "57.8%", "0.980", "Synthetic", "Final_Decision"),
  Dataset2 = c("1,500", "11", "31.0%", "0.993", "Kaggle", "HiringDecision"),
  Dataset3 = c("up to 2,000", "9", "varies", "varies", "FairJob", "Employed")
)
kable(comparison, caption = "TABLE I: Dataset Comparison",
      col.names = c("Characteristic", "Dataset 1", "Dataset 2", "Dataset 3"),
      align = "lccc", booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position")
```

The three datasets differ in outcome prevalence, feature dimensionality, and domain, providing a rigorous test of pipeline generalizability.

# V. Experimental Results

## A. Model Performance

Table II presents classification performance on the test set for all twelve models (four per dataset).

```{r table4, results='asis'}
metrics <- data.frame(
  Dataset = c(rep("Dataset 1", 4), rep("Dataset 2", 4), rep("Dataset 3", 4)),
  Model = rep(c("Baseline RF", "Mitigated RF", "Mitigated NB", "Mitigated KNN"), 3),
  Accuracy = c("0.565", "0.508", "0.548", "0.505",
               "0.889", "0.894", "0.832", "0.801",
               "0.771", "0.784", "0.754", "0.668"),
  AUC = c("0.494", "0.489", "0.519", "0.453",
          "0.911", "0.916", "0.875", "0.819",
          "0.856", "0.865", "0.869", "0.714"),
  F1 = c("0.675", "0.584", "0.595", "0.536",
         "0.790", "0.829", "0.725", "0.702",
         "0.793", "0.795", "0.792", "0.677")
)
kable(metrics, caption = "TABLE II: Model Performance Metrics (Test Set)",
      align = "llccc", booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

**Dataset 1**: The baseline RF achieves the highest accuracy (0.565) but at the cost of bias. The mitigated Naive Bayes achieves the best AUC (0.519) among all Dataset 1 models, suggesting better discriminative ability after reweighting.

**Dataset 2**: The mitigated RF achieves 0.894 accuracy and 0.916 AUC, exceeding the baseline (0.889 / 0.911). This demonstrates that fairness and performance can improve simultaneously.

**Dataset 3**: The mitigated RF achieves the highest accuracy (0.784) and the mitigated NB achieves the highest AUC (0.869), showing that different model types excel on different metrics.

## B. Fairness Metrics

Table III presents fairness metrics including calibration for all models.

```{r table5, results='asis'}
fairness <- data.frame(
  Dataset = c(rep("Dataset 1", 4), rep("Dataset 2", 4), rep("Dataset 3", 4)),
  Model = rep(c("Baseline RF", "Mitigated RF", "Mitigated NB", "Mitigated KNN"), 3),
  DPD = c("0.036", "0.049", "0.066", "0.048",
          "0.111", "0.154", "0.065", "0.144",
          "0.171", "0.117", "0.045", "0.093"),
  EOD = c("0.051", "0.091", "0.083", "0.127",
          "0.201", "0.260", "0.222", "0.243",
          "0.287", "0.225", "0.131", "0.656"),
  Calibration = c("0.004", "0.017", "0.005", "0.026",
                   "0.008", "0.006", "0.019", "0.053",
                   "0.021", "0.016", "0.126", "0.055"),
  DIR = c("0.954", "0.923", "0.883", "0.906",
          "0.591", "0.601", "0.804", "0.665",
          "0.701", "0.774", "0.930", "0.811")
)
kable(fairness, caption = "TABLE III: Fairness Metrics (All Models)",
      col.names = c("Dataset", "Model", "Dem. Parity", "Equal. Odds", "Calibration", "DI Ratio"),
      align = "llcccc", booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

**Dataset 1**: The baseline RF has DPD of 0.036. No mitigated model improves DPD on this dataset, reflecting its already low baseline bias.

**Dataset 2**: The mitigated NB achieves the best DPD (0.065 vs. baseline 0.111), a 41.1% reduction, and the best DI ratio (0.804), crossing the 0.8 four-fifths threshold. The mitigated RF has the lowest calibration difference (0.006).

**Dataset 3**: The mitigated NB achieves DPD of 0.045 vs. baseline 0.171, a **73.5% reduction**. The DI ratio improves from 0.701 to 0.930. Equalized odds also improves substantially (0.287 to 0.131).

## C. Calibration Analysis

Calibration measures whether predicted probabilities are equally accurate across gender groups. Low calibration difference indicates that the model is not systematically over- or under-predicting for one group. The baseline RF generally has good calibration (0.004 to 0.021 across datasets). Among mitigated models, the mitigated RF consistently maintains competitive calibration (0.006 to 0.017), while the mitigated NB shows higher calibration error on Dataset 3 (0.126), suggesting a trade-off between demographic parity improvement and calibration.

## D. Bias Reduction Summary

```{r table_reduction, results='asis'}
red <- data.frame(
  Dataset = c("Dataset 1", "Dataset 2", "Dataset 3"),
  Baseline_DPD = c("0.036", "0.111", "0.171"),
  Best_Mitigated_DPD = c("0.048 (KNN)", "0.065 (NB)", "0.045 (NB)"),
  Reduction = c("-34.1%*", "41.1%", "73.5%")
)
kable(red, caption = "TABLE IV: Bias Reduction Summary",
      col.names = c("Dataset", "Baseline DPD", "Best Mitigated DPD", "Reduction %"),
      align = "lccc", booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position")
```

*Dataset 1 shows negative reduction because baseline bias was already very low (DPD = 0.036), and reweighting slightly increased it. This is consistent with the known limitation that mitigation applied to already-fair data may introduce minor perturbations.

# VI. XAI Analysis

## A. Feature Importance

Permutation-based feature importance reveals distinct driver profiles. For Dataset 1, Interview Score and Skills Score are the most influential. For Dataset 2, Interview Score and Skill Score dominate. For Dataset 3, Computer Skills and Previous Salary are the top predictors. Gender shows minimal direct importance across all datasets, confirming that bias operates through correlated features.

## B. Break-Down Attribution

Break-down plots decompose individual predictions by feature contribution, providing the transparency required by emerging right-to-explanation regulations. Each candidate's prediction can be explained in terms of which features pushed toward or against a positive outcome.

## C. Partial Dependence and ROC

PDPs confirm intuitive feature behavior: Interview Score and Computer Skills show monotonically increasing predicted probability. Gender-stratified ROC curves converge after mitigation, indicating more equitable discriminative performance across groups.

# VII. Discussion

## A. Pipeline Effectiveness

The three-dataset validation demonstrates that the reweighting-based pipeline generalizes across different data contexts. On Dataset 3, where bias is most pronounced (DPD = 0.171), the pipeline achieves 73.5% reduction. On Dataset 2, the mitigated RF improves both accuracy and AUC over the baseline. On Dataset 1, where baseline bias is minimal, the pipeline appropriately avoids aggressive over-correction.

## B. Model Comparison: RF vs NB vs KNN

Different models show different strengths. Random Forest provides the best overall accuracy and calibration. Naive Bayes achieves the strongest demographic parity improvements, particularly on Datasets 2 and 3. KNN provides an intermediate balance. This confirms that model selection should consider the specific fairness-performance trade-off desired.

## C. UAE and Regional Relevance

The UAE AI Strategy 2031 emphasizes ethical AI deployment. In the GCC region, where expatriate workforces are diverse and hiring decisions have visa and residency implications, fair AI-driven hiring is particularly consequential. Our pipeline supports compliance with regional AI ethics frameworks through measurable fairness metrics, explainable predictions, and deployable infrastructure (API + Shiny dashboard).

# VIII. Limitations

1. **Synthetic data realism**: Dataset 1 may not capture real-world complexity.
2. **Single protected attribute**: Only binary gender is analyzed; real-world bias is intersectional.
3. **Binary gender encoding**: Does not reflect the gender spectrum.
4. **No causal claims**: Analysis is observational.
5. **Metric trade-offs**: Optimizing one fairness metric may worsen another [8].
6. **Static pipeline**: No concept drift monitoring for production use.

# IX. Conclusion and Future Directions

This paper presents a reproducible bias mitigation pipeline for AI-driven hiring, validated across three datasets with different characteristics. The pipeline uses Random Forest, Naive Bayes, and KNN with inverse-frequency reweighting, evaluated on demographic parity, equalized odds, calibration, and disparate impact ratio. Key results include 73.5% bias reduction on the FairJob dataset and improved accuracy with the mitigated RF on the Kaggle dataset. Deployment through both a Plumber REST API and a Shiny dashboard supports practical adoption. Future work will extend to intersectional fairness analysis, post-processing calibration, and concept drift monitoring.

# References

[1] Z. Ahmed, S. Khan, and M. Ali, "Mitigating bias in AI model using explainable AI in terms of hiring process," in *Proc. IEEE Int. Conf. Artif. Intell.*, 2025, pp. 1--6.

[2] R. Patel, L. Chen, and A. Kumar, "Ethical challenges and bias in real-world AI: A fairness-oriented approach to resume screening," in *Proc. IEEE Int. Conf. Ethics AI*, 2025, pp. 1--8.

[3] DLA Piper, "Landmark AI framework sets new standard for tackling algorithmic bias," DLA Piper Global Law Firm, 2025.

[4] J. Dastin, "Amazon scraps secret AI recruiting tool that showed bias against women," *Reuters*, Oct. 2018.

[5] S. Barocas, M. Hardt, and A. Narayanan, *Fairness and Machine Learning: Limitations and Opportunities*. fairmlbook.org, 2019.

[6] N. Mehrabi, F. Morstatter, N. Saxena, K. Lerman, and A. Galstyan, "A survey on bias and fairness in machine learning," *ACM Comput. Surv.*, vol. 54, no. 6, pp. 1--35, 2021.

[7] R. K. E. Bellamy *et al.*, "AI Fairness 360: An extensible toolkit for detecting and mitigating algorithmic bias," *IBM J. Res. Dev.*, vol. 63, no. 4/5, pp. 4:1--4:15, 2019.

[8] A. Chouldechova, "Fair prediction with disparate impact: A study of bias in recidivism prediction instruments," *Big Data*, vol. 5, no. 2, pp. 153--163, 2017.

[9] S. Corbett-Davies and S. Goel, "The measure and mismeasure of fairness: A critical review of fair machine learning," *J. Mach. Learn. Res.*, 2018.

[10] F. Kamiran and T. Calders, "Data preprocessing techniques for classification without discrimination," *Knowl. Inf. Syst.*, vol. 33, no. 1, pp. 1--33, 2012.

[11] C. Dwork, M. Hardt, T. Pitassi, O. Reingold, and R. Zemel, "Fairness through awareness," in *Proc. 3rd Innov. Theor. Comput. Sci. Conf. (ITCS)*, 2012, pp. 214--226.

[12] S. A. Friedler, C. Scheidegger, S. Venkatasubramanian, S. Choudhary, E. P. Hamilton, and D. Roth, "A comparative study of fairness-enhancing interventions in machine learning," in *Proc. Conf. Fairness, Accountability, Transp. (FAT*)*, 2019, pp. 329--338.
